---
title: "Vegmon Script"
output:
  word_document: default
  html_document: default
date: "2024-12-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Overstory, Regen, Canopy Analysis

Code developed by Shelby Mead

# Set-up IMPORTANT

Instructions: 

When running this code for a new park/year/etc, some minor changes will need to be made. 
1. Change read-in file names inside quotation marks
2. If graphs are scaled/cropped incorrectly, you may need to alter yaxis limits manually


This chunk of code loads all needed packages

```{r}
##if you have never downloaded these packages, you must download by removing the # in next 4 lines:
#install.packages("tidyverse")
#install.packages("ggplot2")
#install.packages("dplyr")
#install.packages("stringr")
#install.packages("extrafont")



#Otherwise, only these lines will run
library(tidyverse)
library(ggplot2)
library(dplyr)
library(stringr)

```

## Read In Files

```{r}
Canopy <- read.csv("data//HOSPCanopy07_24.csv") #change file name 
Overstory <- read.csv("data//HOSPOverstoryClean07_24.csv")#change file name 
Regeneration <- read.csv("data//HOSPRegeneration07_24Clean.csv")#change file name 
Species<- read.csv("data//tbl_Species_LU.csv")
```

This chunk creates constants that will be used in the code

```{r}
Overstory <- Overstory %>%
  mutate(Class = case_when(DBH < 15 ~ '1',
                           DBH >= 15 & DBH < 25 ~ '2',
                           DBH >= 25 & DBH < 35 ~ '3',
                           DBH >= 35 & DBH < 45 ~ '4',
                           DBH >= 45 ~ '5'))%>% #makes class groupings
  mutate(Year2 = as.character(Year2))%>%
 filter(AcceptedSpecies != "Viburnum rufidulum") #filters out vibu rufi, no longer classified as overstory

n <- length(unique(Canopy$LocationID))  #sets 'n' as number of sites, if not all sites are measured in all years just enter site number manually 

Year <- unique(Overstory$Year2)
Year <- na.omit(Year)
Site <- 1:7
Class <-1:5
supplement <- crossing(Year, Site, Class) #creates dataframe so all combos of year, site, and class can have a zero value

Current <- max(Year)
First <- min(Year)


theme <- theme(axis.text.x = element_text(color = "grey20", size = 12, angle = 0, hjust = .5, vjust = .5, face = "plain",family     = "sans"),
        axis.text.y = element_text(color = "grey20", size = 12, angle = 0, hjust = 1, vjust = 0, face = "plain", family = "sans"),  
    axis.title.x = element_text(color = "grey20", size = 14, angle = 0, hjust = .5, vjust = 0, face = "plain", family
    = "sans"),
    axis.title.y = element_text(color = "grey20", size = 14, angle = 90, hjust = .5, vjust = .5, face = "plain", family = "sans"),
    legend.text = element_text(size = 12, hjust = .5, face = "plain",family = "sans"),
    legend.title = element_text(size = 14, face = "plain", family = "sans"),
    title = element_text(size = 14, face = "plain", family = "sans"),
    plot.margin = unit(c(0.5,0.5,0.5,0.5), 'cm'))

  
```

# Basal Area

##Live 
This chunk runs BA- live analysis

```{r}

BA_Site_Class <- Overstory %>%
  filter(Condition == "L",
           SpeciesCode != "SNAG") %>% #sort live and dead
  mutate(BA_ha = DBH^2*(0.00007854)/0.1,
         Year2 = as.character(Year2))%>% #calculate basal area per ha
  group_by(Year2, Site, Class)%>%
  summarise(BA_site = sum(BA_ha)) #calculate BA per site by class


BA_Site_Class <- supplement %>%
  mutate(Class = as.character(Class))%>%
  left_join(BA_Site_Class, by = c('Year' = 'Year2', 'Site', 'Class')) %>%
  replace(is.na(.),0) #joining Ba data to dataframe with all combos of year and site

BA_Year_Class <- BA_Site_Class %>%
  group_by(Year, Class)%>%
  summarise(BA_Year_Class = (sum(BA_site)/n), #calculating average BA for each class
            SD = sd(BA_site))%>%  #calculating standard deviation of BA for each class
  mutate(CI_low =  BA_Year_Class - (1.959964*(SD/sqrt(n))), #calculating low limit of Confidence interval
         CI_high = BA_Year_Class + (1.959964*(SD/sqrt(n)))) #calculating high limit of confidence interval
write.csv(BA_Year_Class, 'Tables//BasalAreaClass.csv')
BA_Year_Class

BA_Year <- BA_Site_Class %>%
  group_by(Year, Site) %>%
  summarise(BA_Site = sum(BA_site))%>%
  group_by(Year) %>%
  summarise(BA_Year = (sum(BA_Site)/n), #calculating average BA per year
            SD_BA_Year = sd(BA_Site))%>% #standard deviation of BA each year
  mutate(CIlow =  BA_Year - (1.959964*(SD_BA_Year/sqrt(5))),#low limit of confidence interval
         CIhigh = BA_Year + (1.959964*(SD_BA_Year/sqrt(5)))) #high limit of confidence interval
write.csv(BA_Year, 'Tables//BasalArea.csv')
BA_Year


```

This chunk creates BA-live graphs

```{r}
YearBA <- ggplot(BA_Year)+ #makes a plot using BA_Year dataset we just created
  geom_col(aes(x = Year, y = BA_Year),width= 0.5, fill = 'gray60', color = 'gray30')+ #bar graph and specifications
 labs(y = bquote('Mean Basal Area ' ~ (m^2/ha)))+
  geom_errorbar(aes(x = Year, ymin = CIlow, ymax= CIhigh), width = 0.1, size =0.2)+ #adds error bars
  scale_y_continuous(expand = c(0,0), limits = c(0, 1.05*max(BA_Year$CIhigh)))+
  theme_light()+
  theme
YearBA
ggsave('BasalArea.jpg', dpi = 600, path = "Graphs") #saves basal area graph to Output Graphs

BA_by_class <- ggplot(BA_Year_Class, aes(x = Class, y =BA_Year_Class, fill=Year))+ #makes plot using our BA_Year_Class dataset, color by class
  geom_col(position = 'dodge')+ #position dodge makes columns adjacent
  scale_fill_grey()+ #manual color scale of bars
  labs(y = bquote('Mean Basal Area ' ~ (m^2/ha)))+
  geom_errorbar(aes(x = Class, ymin = ifelse(CI_low>0, CI_low,0), ymax=CI_high), position = position_dodge(0.9), width =0.2, size =0.2)+
  #adds error bars
  scale_y_continuous(expand = c(0,0), limits = c(0, 1.05*max(BA_Year_Class$CI_high)))+
  theme_light()+
  theme
BA_by_class
ggsave('BasalAreaClass.jpg', dpi = 600, path= "Graphs", height =5)#saves basal area by class graph to Output Graphs
```

## Dead

This chunk runs BA-dead analysis

```{r}

BA_Site_Class_D <- Overstory %>%
  filter(Condition == "D") %>% #sort live and dead
  mutate(BA_ha = DBH^2*(0.00007854)/0.1,
         Year2 = as.character(Year2))%>% #calculate basal area per ha
  group_by(Year2, Site, Class)%>%
  summarise(BA_site = sum(BA_ha)) #calculate BA per site by class


BA_Site_Class_D <- supplement %>%
  mutate(Class = as.character(Class))%>%
  left_join(BA_Site_Class_D, by = c('Year' = 'Year2', 'Site', 'Class')) %>%
  replace(is.na(.),0) #joining extra dataframe to data

BA_Year_Class_D <- BA_Site_Class_D %>%
  group_by(Year, Class)%>%
  summarise(BA_Year_Class = (sum(BA_site)/n), #calculating average BA for each class
            SD = sd(BA_site))%>%  #calculating standard deviation of BA for each class
  mutate(CI_low =  BA_Year_Class - (1.959964*(SD/sqrt(n))), #calculating low limit of Confidence interval
         CI_high = BA_Year_Class + (1.959964*(SD/sqrt(n)))) #calculating high limit of confidence interval
write.csv(BA_Year_Class_D, 'Tables//BasalAreaClassD.csv')
BA_Year_Class_D

BA_Year_D <- BA_Site_Class_D %>%
  group_by(Year, Site) %>%
  summarise(BA_Site = sum(BA_site))%>%
  group_by(Year) %>%
  summarise(BA_Year = (sum(BA_Site)/n), #calculating average BA per year
            SD_BA_Year = sd(BA_Site))%>% #standard deviation of BA each year
  mutate(CIlow =  BA_Year - (1.959964*(SD_BA_Year/sqrt(5))),#low limit of confidence interval
         CIhigh = BA_Year + (1.959964*(SD_BA_Year/sqrt(5)))) #high limit of confidence interval
write.csv(BA_Year_D, 'Tables//BasalAreaD.csv')
BA_Year_D


```

This chunk creates BA graphs for dead trees

```{r}
YearBA_D <- ggplot(BA_Year_D)+ #makes a plot using BA_Year dataset we just created
  geom_col(aes(x = Year, y = BA_Year),width= 0.5, fill = 'gray60', color = 'gray30')+ #bar graph and specifications
 labs(y = bquote('Mean Basal Area ' ~ (m^2/ha)))+
  geom_errorbar(aes(x = Year, ymin = CIlow, ymax= CIhigh), width = 0.1, size =0.2)+ #adds error bars
  scale_y_continuous(expand = c(0,0), limits = c(0, 1.05*max(BA_Year$CIhigh)))+
  theme_light()+
  theme
YearBA_D
ggsave('BasalAreaD.jpg', dpi = 600, path = "Graphs") #saves basal area graph to Output Graphs

BA_by_class_D <- ggplot(BA_Year_Class_D, aes(x = Class, y =BA_Year_Class, fill=Year))+ #makes plot using our BA_Year_Class dataset, color by class
  geom_col(position = 'dodge')+ #position dodge makes columns adjacent
  scale_fill_grey()+ #manual color scale of bars
  labs(y = bquote('Mean Basal Area ' ~ (m^2/ha)))+
  geom_errorbar(aes(x = Class, ymin = ifelse(CI_low>0, CI_low,0), ymax=CI_high), position = position_dodge(0.9), width =0.2, size =0.2)+
  scale_y_continuous(expand = c(0,0), limits = c(0, 1.05*max(BA_Year_Class$CI_high)))+
  theme_light()+
  #adds error bars
  theme
BA_by_class_D
ggsave('BasalAreaClassD.jpg', dpi = 600, path= "Graphs", height =5)#saves basal area by class graph to Output Graphs
```

# Density

## Live

This chunk runs the density analysis

```{r}
Density <- Overstory %>%
  filter(Condition == 'L', SpeciesCode != "SNAG")%>% #filters out dead trees and snags
  group_by(Year2, Site, Class)%>%
  summarise(n())%>%#count number of stems in each density class for each year, site
  rename(Year = Year2,
         Count = 'n()')%>%
  mutate(Year = as.character(Year)) #making Year value character to join with other tables

Density <- supplement%>%
  mutate(Class = as.character(Class))%>% #Join stem data with supplement table made earlier to 
  left_join(Density, by = c('Year','Site','Class'))%>%
  replace(is.na(.),0)%>%
  mutate(AreaDensity = Count/0.1) #convert to ha
  
Density_Year_Class <- Density %>%
  group_by(Year, Class)%>%
  summarise(StemDensity = mean(AreaDensity), #Mean density by year, class
            SD= sd(AreaDensity))%>% #standard deviation by year, class
   mutate(low = StemDensity - (1.959964*(SD/sqrt(n))),#calculating low limit of Confidence interval
          high = StemDensity + (1.969964*(SD/sqrt(n))))#calculating high limit of Confidence interval
write.csv(Density_Year_Class, 'Tables//DensityClass.csv')
Density_Year_Class

Density_Year <- Density %>%
  group_by(Year,Site)%>%
  summarise(SiteDensity = sum(AreaDensity))%>%
  group_by(Year)%>%
  summarise(StemDensity_Year = (sum(SiteDensity)/n), #calculate mean density by year
            SD = sd(SiteDensity)) %>%#calculate sd by year
  mutate(low = StemDensity_Year - (1.959964*(SD/sqrt(5))),#calculating low limit of Confidence interval
         high = StemDensity_Year + (1.959964*(SD/sqrt(5))))#calculating high limit of Confidence interval
write.csv(Density_Year, 'Tables//Density.csv')
Density_Year
```

This chunk makes density graphs

```{r}
YearDensity <- ggplot(Density_Year, aes(x = Year, y = StemDensity_Year))+ #make ggplot of average density each year, using Density_Year table we just made
  geom_col(fill = 'gray60', width = 0.5, color = 'gray30')+ # setting bar specifications
  ylab('Mean Stem Density (stem/ha)')+ #labeling y axis
  geom_errorbar(aes(x = Year, ymin = low, ymax= high), width = 0.1, size =0.2)+ #adding error bars
  scale_y_continuous(expand = c(0,0), limits = c(0, 1.05*max(Density_Year$high)))+
  theme_light()+
  theme
YearDensity
ggsave('Density.jpg', dpi = 600, path = "Graphs")

ClassDensity <- ggplot(Density_Year_Class, aes(x = Class, y = StemDensity, fill = Year))+ #making ggplot of average density in each size class (1-5) from Density_Year_Class table
  geom_col(position = 'dodge')+ #makes class bars adjacent on x axis
  scale_fill_grey()+ #manually selecting color scheme of bars
  ylab('Mean Stem Density (stems/ha)')+ #labeling y axis
  geom_errorbar(aes(x = Class, ymin =  ifelse(low>0, low,0),  ymax=high), position = position_dodge(0.9), width =0.2, size =0.2)+ #adding error bars 
  scale_y_continuous(expand = c(0,0), limits = c(0, 1.05*max(Density_Year_Class$high)))+
  theme_light()+
  theme

ClassDensity
ggsave('DensityClass.jpg', dpi = 600, path = "Graphs")
```

## Dead

This chunk runs the density analysis for dead trees

```{r}
Density_D <- Overstory %>%
  filter(Condition == 'D')%>% #filters only Dead trees
  group_by(Year2, Site, Class)%>%
  summarise(n())%>%#count number of stems in each density class for each year, site
  rename(Year = Year2,
         Count = 'n()')%>%
  mutate(Year = as.character(Year)) #making Year value character to join with other tables

Density_D <- supplement%>%
  mutate(Class = as.character(Class))%>% #Join stem data with supplement table made earlier to 
  left_join(Density_D, by = c('Year','Site','Class'))%>%
  replace(is.na(.),0)%>%
  mutate(AreaDensity = Count/0.1) #convert to ha
  
Density_Year_Class_D <- Density_D %>%
  group_by(Year, Class)%>%
  summarise(StemDensity = mean(AreaDensity), #Mean density by year, class
            SD= sd(AreaDensity))%>% #standard deviation by year, class
   mutate(low = StemDensity - (1.959964*(SD/sqrt(n))),#calculating low limit of Confidence interval
          high = StemDensity + (1.969964*(SD/sqrt(n))))#calculating high limit of Confidence interval
write.csv(Density_Year_Class_D, 'Tables//DensityClassD.csv')
Density_Year_Class_D

Density_Year_D <- Density_D %>%
  group_by(Year,Site)%>%
  summarise(SiteDensity = sum(AreaDensity))%>%
  group_by(Year)%>%
  summarise(StemDensity_Year = (sum(SiteDensity)/n), #calculate mean density by year
            SD = sd(SiteDensity)) %>%#calculate sd by year
  mutate(low = StemDensity_Year - (1.959964*(SD/sqrt(5))),#calculating low limit of Confidence interval
         high = StemDensity_Year + (1.959964*(SD/sqrt(5))))#calculating high limit of Confidence interval
write.csv(Density_Year_D, 'Tables//DensityD.csv')
Density_Year_D
```

This chunk makes density graphs for dead trees

```{r}
YearDensity_D <- ggplot(Density_Year_D, aes(x = Year, y = StemDensity_Year))+ #make ggplot of average density each year, using Density_Year table we just made
  geom_col(fill = 'gray60', width = 0.5, color = 'gray30')+ # setting bar specifications
  ylab('Mean Stem Density (stem/ha)')+ #labeling y axis
  geom_errorbar(aes(x = Year, ymin = low, ymax= high), width = 0.1, size =0.2)+ #adding error bars
  scale_y_continuous(expand = c(0,0), limits = c(0, 1.05*max(Density_Year$high)))+
  theme_light()+
  theme+
  coord_cartesian(ylim = c(0,1000),clip='off')
YearDensity_D
ggsave('DensityD.jpg', dpi = 600, path = "Graphs")

ClassDensity_D <- ggplot(Density_Year_Class_D, aes(x = Class, y = StemDensity, fill = Year))+ #making ggplot of average density in each size class (1-5) from Density_Year_Class table
  geom_col(position = 'dodge')+ #makes class bars adjacent on x axis
  scale_fill_grey()+ #manually selecting color scheme of bars
  ylab('Mean Stem Density (stems/ha)')+ #labeling y axis
  geom_errorbar(aes(x = Class, ymin = ifelse(low>0, low, 0), ymax=high), position = position_dodge(0.9), width =0.2, size =0.2)+
  scale_y_continuous(expand = c(0,0), limits = c(0, 1.05*max(Density_Year_Class$high)))+#adding error bars 
  theme_light()+
  theme

ClassDensity_D
ggsave('DensityClassD.jpg', dpi = 600, path = "Graphs")
```

#Species-Wise Analysis

## Live

```{r}
## Calculating species change by BA
Species_BA_Site <- Overstory %>%
  filter(Condition == "L")%>% #filter for only Live trees
  mutate(BA_ha = DBH^2*(0.00007854/0.1)) %>% #convert DBH to BA scaled to hectare
  group_by(Year2, Site, AcceptedSpecies)%>%
  summarise(BA_site = sum(BA_ha))%>% # summarise basal area for each species in a site
  na.omit()#calculate BA per site by species
  
Species_BA <- Species_BA_Site %>% 
  group_by(Year2, AcceptedSpecies)%>% 
  summarise(MeanBA = sum(BA_site/n)) #sumarise basal area mean across sites



PivotSpecies_BA <- Species_BA %>%
  filter(Year2 == Current)%>% #filters for most recent year
  full_join((subset(Species_BA, Year2 == First)), by = "AcceptedSpecies")%>% #filters for first year
  mutate(Year2.x = Current, Year2.y = First, #fills in NA for year
         across(3:5, ~replace_na(.,0)), #replaces NA values with 0
         Change = MeanBA.x - MeanBA.y)%>% #calculates change between first and most recent year
  select(AcceptedSpecies, Change) %>%
  right_join(Species_BA, by = "AcceptedSpecies")%>% #joins with entire table
  arrange(Year2)%>% #Arrange by year to make logical columns
  pivot_wider(names_from = Year2, values_from = MeanBA)%>% #pivot wider to create table
  arrange(AcceptedSpecies) %>% #alphabatize rows
  mutate(across(where(is.double), ~replace_na(.,0)))%>% #replace NA with 0
  relocate(Change, .after = last_col())%>% #relocate change column to after year columns
  mutate(Direction = if_else(Change>0, "Increase", "Decrease")) #designation direction of change for graphs
write.csv(PivotSpecies_BA, "Tables\\SpeciesBA.csv")
  



## Calculating species change by Density
Species_Density_Site <- Overstory %>%
  filter(Condition == "L")%>%
  na.omit()%>% #removing NA values
  group_by(Year2, Site, AcceptedSpecies)%>%
  summarise(Stem_ha = n()/0.1) #convert number of stems to stems/ha

Species_Density <- Species_Density_Site %>%
  group_by(Year2, AcceptedSpecies)%>%
  summarise(MeanDensity = sum(Stem_ha)/n) #calculates year mean density for each species


PivotSpecies_Density <- Species_Density %>%
  filter(Year2 == Current)%>% #filters for most recent year
  full_join((subset(Species_Density, Year2 == First)), by = "AcceptedSpecies")%>% #filters for first year
  mutate(Year2.x = Current, Year2.y = First, #fills in NA for year
         across(3:5, ~replace_na(.,0)), #replaces NA values with 0
         Change = MeanDensity.x - MeanDensity.y)%>% #calculates change between first and most recent year
  select(AcceptedSpecies, Change) %>%
  right_join(Species_Density, by = "AcceptedSpecies")%>% #joins with entire table
  arrange(Year2)%>% #Arrange by year to make logical columns
  pivot_wider(names_from = Year2, values_from = MeanDensity)%>% #pivot wider to create table
  arrange(AcceptedSpecies) %>% #alphabatize rows
  mutate(across(where(is.double), ~replace_na(.,0)))%>% #replace NA with 0
  relocate(Change, .after = last_col())%>% #relocate change column to after year columns
  mutate(Direction = if_else(Change>0, "Increase", "Decrease")) #designation direction of change for graphs
write.csv(PivotSpecies_Density, 'Tables//SpeciesDensity.csv')
PivotSpecies_Density
```

## Occurence table
```{r}

Overstory %>%
  filter(Year2 == Current)%>%
  group_by(AcceptedSpecies)%>%
  mutate(PercentOccurence = (length(unique(Site))/n*100))%>%
  select(AcceptedSpecies, PercentOccurence)%>%
  distinct()%>%
  right_join(as_data_frame(unique(Overstory$AcceptedSpecies)), by = c("AcceptedSpecies" = "value"))%>%
  mutate(PercentOccurence = ifelse(is.na(PercentOccurence), 0, PercentOccurence))%>%
  arrange(AcceptedSpecies)%>%
  left_join(Species, by = "AcceptedSpecies")%>%
  rename(CommonName = USDA_CName)%>%
  select(AcceptedSpecies, CommonName, PercentOccurence)%>%
  write.csv("Tables\\PercentOccurence.csv", row.names = F)

```

This chunk creates change charts by species name

```{r}
## Subtraction change in BA
ggplot(PivotSpecies_BA, aes(y = reorder(AcceptedSpecies, desc(AcceptedSpecies)), x = Change, fill = Direction))+ # sets ggplot to show species change in density
  geom_col(width = 0.4)+ #makes the ggplot a barplot
  xlab('Change in Mean Basal Area')+ #x axis label
  ylab('Species Name')+ #y axis label
  scale_fill_manual(breaks = c('Decrease', 'Increase'),
                    values = c('#FDB338', '#025196'))+ #manually setting color scale
  scale_x_continuous(breaks = scales::pretty_breaks(n=12))+
  theme_light()+
  theme+
  theme(axis.text.y = element_text(face = "italic"),
        axis.text.x = element_text(angle = 90))+
  geom_vline(xintercept = 0)
ggsave('SpBAchange.jpg', dpi = 600, height = 8, width = 8, path = "Graphs")

## subtraction change in Density
ggplot(PivotSpecies_Density, aes(y = reorder(AcceptedSpecies, desc(AcceptedSpecies)), x = Change, fill = Direction))+ # sets ggplot to show species change in density
  geom_col(width = 0.4)+ #makes the ggplot a barplot
  xlab('Change in Mean Stem Density (stems/ha)')+ #x axis label
  ylab('Species Name')+ #y axis label
  scale_fill_manual(breaks = c('Decrease', 'Increase'),
                    values = c('#FDB338', '#025196'))+ #manually setting color scale
  scale_x_continuous(breaks = scales::pretty_breaks(n=12))+
  theme_light()+
  theme+
  theme(axis.text.y = element_text(face = "italic"),axis.text.x = element_text(angle = 90))+
  geom_vline(xintercept = 0)
ggsave('SpDensityChange.jpg', dpi = 600, height = 8, width = 8, path = "Graphs")
```

# Canopy

This chunk of code makes table

```{r}

Canopy1 <- Canopy %>%
  mutate(SumOfValue.1 = na_if(SumOfValue.1, -1), SumOfValue.2 = na_if(SumOfValue.2, -1), SumOfValue.3 = na_if(SumOfValue.3, -1),SumOfValue.4 = na_if(SumOfValue.4, -1))%>% #make -1 values NA
  mutate(C1 = SumOfValue.1 * 1.04, C2 = SumOfValue.2 * 1.04, C3 = SumOfValue.3*1.04, C4 = SumOfValue.4*1.04) %>% #multiply by 1.04
  mutate(PlotMean = (C1+C2+C3+C4)/4) %>% #Calculate Plot mean (n=4)
  group_by(Year, LocationID) %>%
  summarize (SiteMean= mean(PlotMean), SiteSD= sd(PlotMean)) %>% #calculate sitemean and standard deviation
  mutate(Site95high = SiteMean + (1.959964*(SiteSD/sqrt(n))), #calculate high CI interval
         Site95low = SiteMean -  (1.959964*(SiteSD/sqrt(n))))%>% #calculate low CI interval
  mutate(Site = str_extract(LocationID, "\\d+"), #Extract site number from LocationID
         Year= as.character(Year)) 


Canopy2<- Canopy1 %>%
  group_by(Year) %>%
  summarize (YearMean= mean(SiteMean), YearSD= sd(SiteMean)) %>% #calculate park mean
  mutate(CI95high= YearMean +(1.959964*(YearSD/sqrt(n)))) %>%  #Calculate higher limit of 95% Confidence Interval
  mutate(CI95low= YearMean - (1.959964*(YearSD/sqrt(n)))) #calculate lower limit of confidence interval


write.csv(Canopy2, 'Tables//CanopyMean.csv')
Canopy2

```

This chunk of code makes canopy graph

```{r}

Mean_Canopy <- ggplot(Canopy2)+ ## Makes ggplot for year canopy mean
  geom_col(aes(x = Year, y = YearMean), width= 0.5, fill = 'gray60', color = 'gray30')+ ## makes the ggplot a bar graph seperated by year
  ylab('Mean Canopy Cover (%)')+ # y axis label
  geom_errorbar(aes(x = Year, ymin = CI95low, ymax=CI95high), width = 0.1, size =0.2)+ #adding error bars
  scale_y_continuous(expand = c(0,0), limits = c(0, 1.05*max(Canopy2$CI95high)))+
  theme_light()+
  theme
Mean_Canopy
ggsave("CanopyMean.jpg", dpi = 600, path = "Graphs", height = 4)


Canopy_by_site<- ggplot(Canopy1, aes(x = Site, y =SiteMean, fill=Year))+ # makes ggplot of canopy with color indicating site
  geom_col(position = 'dodge')+
   scale_fill_grey()+
  ylab('Mean Canopy Cover (%)')+
  geom_errorbar(aes(x = Site, ymin = Site95low, ymax=Site95high), position = position_dodge(0.9), width =0.2, size =0.2)+
  scale_y_continuous(expand = c(0,0), limits = c(0, 1.05*max(Canopy1$Site95high)))+
  theme_light()+
  theme
Canopy_by_site
ggsave("CanopySite.jpg", dpi = 600, path = "Graphs", height = 4)
```


# Regeneration

This chunk of code calculates regeneration density by  Class

```{r}

Type_Regen <- Regeneration %>%
  rename(Seedling = SumOfSeedling, SmallSapling = SumOfSmallSapling, LargeSapling = SumOfLargeSapling)%>%
  select(Year, LocationID, AcceptedSpecies, Seedling, SmallSapling, LargeSapling)%>%
  pivot_longer(cols = c(Seedling, SmallSapling, LargeSapling), names_to = "Class", values_to = "Count")%>%
  group_by(Year,LocationID, Class)%>%
  summarise(Sum = sum(Count))%>% #Sum all species for a site/year combination
  mutate(Sum_ha = Sum/0.01,
         Class = ifelse(Class == "LargeSapling", "Large Sapling", ifelse(Class == "SmallSapling", "Small Sapling", "Seedling")))%>% #scale by a factor of 10
  group_by(Year, Class)%>% 
  summarise(MeanDensity = (sum(Sum_ha)/n), #calculate mean park density
            SD = sd(Sum_ha))%>% #calculate SD
  mutate(low = MeanDensity - (1.959964*(SD/sqrt(5))), #calculate low end of 95 conf interval
         high = MeanDensity + (1.959964*(SD/sqrt(5))),#calculate low end of 95 conf interval
         Year = as.character(Year))

write.csv(Type_Regen, "Tables//RegenType.csv")
Type_Regen

```

This chunk makes a graph of regeneration density by class 

```{r}
TypeRegen_graph <-Type_Regen %>%
  arrange(MeanDensity)%>%
  mutate(Class = factor(Class, levels=c("Seedling", "Small Sapling",  "Large Sapling")))%>% #manually setting order based on size, not alphabetized
  ggplot(aes(x = Class, y =MeanDensity, fill=Year))+
  geom_col(position = "dodge")+ #sets bars next to each other
  geom_errorbar(aes(x = Class, ymin = low, ymax= high), width = 0.2, position = position_dodge(0.9), , size =0.2)+ #add error bar
  ylab('Mean Density (stems/ha)')+
  scale_fill_grey()+
  scale_y_continuous(expand = c(0,0), limits = c(0, 1.05*max(Type_Regen$high)))+
  theme_light()+
  theme

TypeRegen_graph
ggsave('RegenType.jpg', path = "Graphs", dpi= 600)
  
```

Regen by species analysis code

```{r}


#this creates the dataframe used to generate graphs

RegenSpecies1 <- Regeneration %>%
  rename(Seedling = SumOfSeedling, SmallSapling = SumOfSmallSapling, LargeSapling = SumOfLargeSapling)%>%
  select(Year, LocationID, AcceptedSpecies, Seedling, SmallSapling, LargeSapling) %>%
  pivot_longer(cols = c(Seedling, SmallSapling, LargeSapling), names_to = "Class", values_to = "Count")%>% #making each class its own row
  mutate(Density_ha = Count/0.01, #convert to stem/ha
         Year = as.character(Year))%>% #Year as character
  select(!Count)%>%
  group_by(Year, AcceptedSpecies, Class)%>% #sort by species each year
  summarise(Year_Density = (sum(Density_ha)/n)) #summarise the year mean density class

RegenSpecies <- RegenSpecies1 %>%
  filter(Year == Current)%>%#filters for most recent year
  full_join((subset(RegenSpecies1, Year == min(Year))), by = c("AcceptedSpecies", "Class")) %>% #filters for first year
  mutate(Year.x = max(Year), Year.y = min(Year), #fills in NA for year
         across(3:5, ~replace_na(.,0)), #replaces NA values with 0
         Change = Year_Density.x - Year_Density.y)%>% #calculates change
  right_join(RegenSpecies1, by = c("AcceptedSpecies", "Class"))%>%
  group_by(AcceptedSpecies, Class)%>%
  mutate(check = sum(Year_Density))%>%
  filter(check != 0)%>%
  mutate(Change = ifelse(is.na(Change), 0, Change),
         Direction = if_else(Change>0, "Increase", "Decrease"))%>%
  select(AcceptedSpecies, Class, Change, Direction)%>%
  distinct()%>%
  arrange(AcceptedSpecies)
  


#this creates the wider dataframe to export
ClassChanges <- RegenSpecies %>%
  select(!Direction)%>%
  pivot_wider(names_from = Class, values_from = Change)%>%
  rename(Change_Seedling = Seedling,
         Change_SmallSapling = SmallSapling,
         Change_LargeSapling = LargeSapling)

Regen_Species_Table <- RegenSpecies1 %>%
  pivot_wider(names_from = Class, values_from = Year_Density)%>%
  pivot_wider(names_from = Year, values_from = c(Seedling, SmallSapling, LargeSapling)) %>%
  mutate_if(is.numeric, ~replace_na(.,0))%>%
  arrange(AcceptedSpecies)%>%
  left_join(ClassChanges, by = "AcceptedSpecies")
  
write.csv(Regen_Species_Table, "Tables\\RegenSpType.csv", row.names = F)

Regen_Species_Table
  
```
This chunk makes change graphs for regeneration by species, each class is graphed alone
```{r}

## Seedling
RegenSpecies %>%
  filter(Class == 'Seedling') %>% #filter for seedlings
  ggplot(aes(y = reorder(AcceptedSpecies, desc(AcceptedSpecies)), x = Change, fill = Direction))+ #ggplot y axis barplot with colors based on increase or decrease
  geom_col(width = 0.4)+ #width of columns
  xlab('Change in Mean Density (Stems/ha)')+
  ylab('Species Name')+
  labs(title = 'Seedlings')+
  scale_fill_manual(breaks = c('Decrease','Increase'),
                    values = c('#FDB338', '#025196'))+ #sets color for increase/decrease
  scale_x_continuous(breaks = scales::pretty_breaks(n=15))+ #changes number of ticks on x axis
  theme_light()+
  theme+
  theme(axis.text.y = element_text(face = "italic"), #makes species names italic
        axis.text.x = element_text(angle = 90))+
  geom_vline(xintercept = 0)
ggsave('RegenSpSeedling.jpg', path = "Graphs", dpi= 600, height = 6)
  
## Small Sapling
RegenSpecies %>%
  filter(Class == 'SmallSapling') %>% #filter for small saplings
  ggplot(aes(y = reorder(AcceptedSpecies, desc(AcceptedSpecies)), x = Change, fill = Direction))+#ggplot y axis barplot with colors based on increase or decrease
  geom_col(width = 0.4)+#width of columns
  xlab('Change in Mean Density (Stems/ha)')+
  ylab('Species Name')+
  labs(title = 'Small Saplings')+
  scale_fill_manual(breaks = c('Decrease','Increase'),
                    values = c('#FDB338', '#025196'))+#sets color for increase/decrease
  scale_x_continuous(breaks = scales::pretty_breaks(n=15))+#changes number of ticks on x axis
  theme_light()+
  theme+
  theme(axis.text.y = element_text(face = "italic"),#makes species names italic
        axis.text.x = element_text(angle = 90))+
  geom_vline(xintercept = 0)
ggsave('RegenSpSmallSap.jpg', path = "Graphs", dpi= 600, height = 6)

## Large Sapling
RegenSpecies %>%
  filter(Class == 'LargeSapling') %>% #filter for large saplings
  ggplot(aes(y = reorder(AcceptedSpecies, desc(AcceptedSpecies)), x = Change, fill = Direction))+#ggplot y axis barplot with colors based on increase or decrease
  geom_col(width = 0.4)+#width of columns
  xlab('Change in Mean Density (Stems/ha)')+
  ylab('Species Name')+
  labs(title = 'Large Saplings')+
  scale_fill_manual(breaks = c(breaks = c('Decrease','Increase')),
                    values = c('#FDB338', '#025196'))+#sets color for increase/decrease
  theme_light()+
  theme+
  theme(axis.text.y = element_text(face = "italic"),axis.text.x = element_text(angle = 90))+#makes species names italic
  geom_vline(xintercept = 0)
ggsave('RegenSpLargeSap.jpg', path = "Graphs", dpi= 600, height = 6)

```
